EuPPollNet: A European database of plant-pollinator networks

### 2024-05-26 Manuscript/EuPPollNet_Ms.qmd
edits

Although Europe contains a much larger number of flowering plants than pollinator species (~4 to 1 ratio), the observed number of pollinator species almost doubled the one of plants in the database. This could be explained by the fact that all networks are phytocentric, resulting in sampling bias towards pollinator species [@jordano2016; @vizentin2018]. While animal-centred sampling is likely to increase the plant:pollinator species ratio [e.g., @encinas2023], the spatial scale and environmental context of the sampled communities will also influence their observed diversity, especially given the ability to move of pollinators and the sessile nature of plants. As expected [@jordano1987], we found that network connectance decreased exponentially with the number of species. Overall, latitude did not explain major changes in residual connectance or normalised nestedness of the plant-pollinator networks. Consistent with @olesen2002, residual connectance was lower at higher latitudes, while normalised nestedness, which is negatively correlated with residual connectance, increased towards higher latitudes. Given their correlation, the potential ecological explanations that can help understanding these structural patterns can be similar. For instance, networks at lower latitudes in Europe are exposed to higher temperatures and are bee-dominated, which can result in higher visitation rates [@arroyo1985; @classen2015; @herrera2019]; and, the level of pollinator generalization seems to be higher at lower latitudes [@schleuning2012]. Therefore, these different factors can increase the possible number of connections that can be established between plants and pollinators for a given network size, resulting in more connected networks but less nested ones in comparison to networks at higher latitudes. In addition, while one third of networks were more nested than expected by chance, two thirds of them did not show statistical difference with null expectations. This goes against the common view that most plant-pollinator networks are nested [@bascompte2003], and supports the idea that despite a strong tendency towards it, being nested is not the rule fo this large set of plant-pollinator networks.

Despite this database covering a wide range of habitats across `r country_number` countries, it contains geographical biases that can impact our understanding of plant-pollinator communities [@hughes2021]. For instance, most plant-pollinator networks belongs to central Europe, while Eastern Europe and the Mediterranean region are under-represented. This is consistent with previous studies which also report lack of plant-pollinator data for those regions [REFs], highlighting that this database reflects existing patterns in data availability despite the absence of a systematic search for studies. This is especially relevant as Eastern Europe is expected to continue developing and use their vast natural resources [@bennett2018], and the Mediterranean region is likely to be severely impacted by climate change [@jaworski2022; @duchenne2020; @pareja2023]. These areas are well known for their rich pollinator diversity, and their under-representation is likely contributing to the low taxonomic coverage of this database.

>>>>

Although Europe contains a much larger number of flowering plants than pollinator species (~4 to 1 ratio), the observed number of pollinator species almost doubled the one of plants in the database. This could be explained by the fact that all networks are phytocentric, resulting in sampling bias towards pollinator species [@jordano2016; @vizentin2018]. While animal-centred sampling is likely to increase the plant:pollinator species ratio [e.g., @encinas2023], the spatial scale and environmental context of the sampled communities will also influence their observed diversity, especially given the ability to move of pollinators and the sessile nature of plants. As expected [@jordano1987], we found that network connectance decreased exponentially with the number of species. Overall, latitude did not explain major changes in corrected connectance or normalised nestedness. However, consistent with @olesen2002, corrected connectance was lower at higher latitudes, while normalised nestedness, which is negatively correlated with corrected connectance, increased towards higher latitudes. Thus, the potential ecological explanations that can help understanding these associations can be similar. For instance, networks at lower latitudes in Europe are exposed to higher temperatures and are bee-dominated, which can result in higher visitation rates [@arroyo1985; @classen2015; @herrera2019]. Further, the level of pollinator generalization seems to be higher at lower latitudes [@schleuning2012]. Therefore, these different factors can increase the possible number of connections that can be established between plants and pollinators for a given network size, resulting in more connected networks but less nested ones in comparison to networks at higher latitudes. In addition, while one third of networks were more nested than expected by chance, two thirds of them did not show statistical difference with null expectations. This goes against the common view that most plant-pollinator networks are nested [@bascompte2003], and supports the idea that despite a strong tendency towards it, being nested is not the rule fo this large set of plant-pollinator networks.

Despite this database covering a wide range of habitats across `r country_number` countries, it contains geographical biases that can impact our understanding of plant-pollinator communities [@hughes2021]. For instance, most plant-pollinator networks belongs to central Europe, while Eastern Europe and the Mediterranean region are under-represented. This is consistent with previous studies which also report lack of plant-pollinator data for those regions [REFs], highlighting that this database reflects existing patterns in data availability despite the absence of a systematic search for studies. This is especially relevant as Eastern Europe is expected to continue developing and use their vast natural resources [@bennett2018], and the Mediterranean region is likely to be severely impacted by climate change [@jaworski2022; @duchenne2020; @pareja2023]. These areas are well known for their rich pollinator diversity [REFs], and their under-representation is likely contributing to the low taxonomic coverage of this database at the European level.
